# Inverse Transform Method (ITM)

Two requirements to use the Inverse Transform Method (ITM) to generate from a random variable $X$ with cdf $F_X(x)$:

1.  a uniform random number generator `runif( )`

2.  the inverse cdf, $F^{−1}_X(p)$

::: callout-tip
### Steps

1.  Compute the inverse of $F_X(p):F^{−1}_X(p)$

2.  Write an `R` function that computes $F^{−1}_X(p)$

3.  Draw $u \sim U(0, 1)$ and set $F^{−1}_X(u)$
:::

#### Note

-   for a discrete distribution computing the CDF is not difficult: add up the individual probabilities for the points of the distribution

-   CDF for continuous dist. integrate the PDF

### Example: exponential dist

**Step 1**: find the inverse $F_X(p)$, the cdf is: $F_x = 1- exp(-\lambda x)$ for inverse solve:

$$
F_x() = p  \\
F_x() = 1- e^{(-\lambda x)} = p \\
e^{(-\lambda x)} = 1 - p \\
ln(e^{(-\lambda x)}) = ln(1-p) \\
-\lambda x = ln(1-p) \\
x = - \frac{1}{\lambda} \cdot ln(1-p)
$$

**Step 2 &3:**

```{r}
set.seed(1234)
random.exp <- function(n, lam) {
  u <- runif(n)
  - log(1 - u) / lam
}

x <- random.exp(1e6, 3)
```

```{r}
#| label: fig-exp.ex
#| fig-cap: "Mean = `r mean(x)` and variance `r var(x)` E(X) if X~exp is 1/lam and var(x) should be 1/(lam^2)"
hist(x, 50, prob = T)
curve(3 * exp(-3 * x), add = TRUE, col = 'red', lwd = 2)

```

### Lecture 1: Exercise 2

#### Step 1: pdf, cdf and expected value

X \~ continuous random variable with density $f_X(x) = x^3$ for 0\<=x\<=k

1.  find k such that $f_X(x)$ is a pdf

-   integrate the pdf

$$
\int_{0}^{k} x^3 dx \\
\left[ \frac{x^4}{4} \right]_0^k \\ 
\frac{k^4}{4}- 0 = 1 \\
k^4 = 1*4 \\
k = \sqrt{2}
$$

2.  Compute the cdf FX(x)

The cdf is the integral of the pdf over it's range

$$
F_X(x)= \int_{0}^{x} f_X(s) ds \\
= \int_{0}^{x} s^3 ds \\
= \frac{x^4}{4}, \text{ when } 0<=x<=k = \sqrt{2}   \\ 
$$

Therefore the cdf is given by:

$$
F_X(x) = 
\begin{cases} 
0, & \text{if  } x < 0; \\
\frac{1}{4}x^4, & \text{if  } 0 \leq x \leq \sqrt{2}; \\
1, & \text{if  } x > \sqrt{2}.
\end{cases}
$$

3.  Compute E(X)

$$

E(X) = \int_{-\infty}^{\infty} x f_X(x) dx \\

E(X) = \int{0}^{k} x \cdot x^3 dx = \int{0}^{k} x^4 dx\\

E(X) = \left[ \frac{x^5}{5} \right]_0^k \\

E(X) = \left[ \frac{x^5}{5} \right]_0^{\sqrt 2} \\

E(X) = \left[ \frac{(\sqrt 2)^5}{5} \right] - 0 \\

E(X) = \frac{1}{5}*(*\sqrt 2)^4 (\sqrt 2) \\

E(X) = \frac{4}{5}(\sqrt 2)

$$

3.2 Compute Var(X) $Var(X) = E[X^2] -E[X]^2$

Step 1 
$$ 

E(X^2) = \int{0}^{k} x^2\cdot x^3 dx = \int{0}^{k} x^5 dx\\

= \left[ \frac{x^6}{6} \right]_0^k \\

= \left[ \frac{(\sqrt 2)^6}{6} \right] - 0 \\

= \frac{8}{6} = \frac{4}{3} 
$$

Step 2 $[E(X)]^2 =  (\frac{4}{5}(\sqrt 2))^2$


$$
Var(X) = \frac{4}{3} - (\frac{4}{5}(\sqrt 2))^2 = \frac{4}{75}
$$

#### Step 2: application of the ITM

Apply the ITM to generate (pseudo)-random numbers from this random variable:

**Step 1** Compute the inverse cdf $F^{−1}_X (x)$ 

For $0<=x<=k = \sqrt{2}$: cdf: $F_X^{-1}(x) = \frac{x^4}{4}$

Set $F_X^{-1}(x) = p$ : 

$$ 
\frac{x^4}{4} = p \\ 
x^4 = 4p \\ x = (4p)^{1/4} \\

F_X^{-1}(p) = (4p)^{1/4}

$$

and (b) write an R function to generate random numbers from X ∼ fX(x) that relies on the ITM.

```{r}
itm_method <- function(n){
  u <- runif(n)
  (4*u)^(1/4)
}

```

#### Step 3: double-check your random number generator

```{r}
set.seed(249124) # 6

X <- itm_method(10000) #7

```

8.  Compare the range of the numbers generated at point (7) to the support of X. What do you notice?

```{r}
summary(X)
```

The support of X is ($0, \sqrt2$) but the range of the sample is (0.11, 1.4).

9.  Compute the mean and variance of the numbers generated at point (7), and compare them to the values of E(X) and V ar(X) that you computed at point (3). Are the numerical results in agreement with your analytical derivations?

```{r}
mean(X); var(X)
```

analytical:

```{r}
4/3
4/75
```

10. Draw a histogram that shows the empirical density of the 10000 random numbers, and compare it to fX(x)

```{r}
hist(X, prob = TRUE, breaks = 50)
curve(x^3, add = TRUE, col = "cornflowerblue")
```

The values generated are close to the curve of $f_X(x)$

# (lecture 2) Discrete R.V.

#### Example:  binomial distribution

```{r}
n = 7; p = 0.4
omega = 0:n
pmf = dbinom(omega, size = n, prob = p) # pmf of # X \sim Bin(n, p)
bin.distr = data.frame(omega, pmf)
set.seed(2)
# how to sample 10 random numbers from X \sim Bin(n, p): 
sample(x = bin.distr$omega, size = 10,
       prob = bin.distr$pmf, replace = T)

x = sample(x = bin.distr$omega, size = 1e5, prob = bin.distr$pmf, replace = T)
bin.distr$rel.freq = as.numeric(table(x)/length(x)) 
round(bin.distr, 5)
```

-   sample(1:k, n, replace = T, prob = c(...)) allows you to generate n independent realizations from a multinomial distribution

-   each single draw from sample( ) as an application of the ITM

#### Example

$\Omega = {1, 2, 3, 4}$, $P(X = k) = \frac{k}{10}$ for $\Omega \in k$

1.  pdf: $P(X = k) = \frac{k}{10}$ Where k can be 1,...4

2.  cdf is defined as $F_X(x) = P(X <= k)$ i.e. the sum of probabilities of all the outcomes of k.

$$
F_X(x) = 
\begin{cases} 
0                                           &  x < 1; \\
\frac{1}{10}                                & 1 \leq x \leq{2} 
                                                \text{ i.e.  }P(X=1); \\
\frac{1}{10}+\frac{2}{10} = \frac{3}{10}    & 2 \leq x \leq{3}; \\
\frac{6}{10}                                & 3 \leq x \leq{4}; \\
1                                           &  x <4; \\

\end{cases}
$$
##### Applying ITM 

-   Draw from a U(0, 1)  
-   Apply the inverse CDF to determine the outcome 

Suppose we generate u1 = 0.55
This u is in the interval of $\frac{3}{10}\leq x < \frac{6}{10}$
The corresponding value of X for this interval is 3
Therefore $F_X^{-1}(0.55) =3$


