# Multiplicative Congruential Random Number Generator (MCRNG)

(from:  Braun and Murdoch's book "A first course in statistical programming with R" 2nd edition (2016): Section 5.2)

## Monte Carlo Simulation 

to approximate the mean $\mu = E(x)$: 
generate m independent and identically distributed (i.i.d.) values of X and use the 
sample mean ($\bar{X} = \frac{\Sigma X_i}{m}$) : good approximation for large m (law of large numbers)
the distribution of $\bar{X}$ can be approximated by a normal distribution the mean 
$\mu$ and variance $\sigma ^2 /m$ -> central limit theorem 


A multiplicative congruential random number generator (MCRNG) is an algorithm 
that can be used to draw numbers from $X \sim U(0, 1)$

m large integer, b is another integer where m>b
$x_0$ integer between 1 and m <- seed

```{r}
m <- 7
b <- 3
x0 <- 2
x1 <- (b*x0)%%m
u1 <- x1/m
x2 <- (b*x1)%%m
u2 <- x2/m

```

formula 
$x_n = bx_{n-1}$(mod m), $ u_n = x_n / m$

```{r}
mcrng <- function(m = 30269, b = 171, x0 = 27218, n = 10){
  X <- numeric(n+1)
  X[1] <- x0
  U <- numeric(n)
  for (i in 1:length(U)) {
    X[i+1]<- (b*X[i])%%m
    U[i]<- X[i+1]/m
  }
  return(U)
}

ex1 <- mcrng(m = 30307, b = 172, n = 20, x0 = 17218)
ex1
```
- choose m so that it is not divizable by b 

ex 4
```{r}
set.seed(19908)
U <- runif(n =1000)
#(a)
(ex4a <- c(mean(U), var(U), sd(U)))
```
(b) Compare your results with the true mean, variance, and standard deviation

mean = median of uniform distribution is 1/2(a+b)
variance: $1/12(b-a)^2$ = 1/12
sd = sqrt(var)

```{r}
0.5
(1/12)
sqrt(1/12) 
```

answer the same to 2d.p

#### ex 5
5 Simulate 10000 independent observations on a uniformly distributed
random variable on the interval [3.7, 5.8].
(a) Estimate the mean, variance, and standard deviation of such a uniform random variable and compare your estimates with the true
values.

```{r}
u <- runif(1000, min = 3.7, max = 5.8)
# true 
(3.7+5.8)/2
# est
mean(u)

## variance
# true 
1/12 * (5.8- 3.7)^2
# est
var(u)

## SD
# true 
sqrt(1/12 * (5.8- 3.7)^2)
# est
sd(u)

```
(b) Estimate the probability that such a random variable is greater than
4.0. Compare with the true value.
```{r}
sum(u>4)/ 1000

(5.8 - 4)/(5.8 - 3.7)

```



### Steps 

-   draw a sample $x$ with replacement from a discrete uniform with $\Omega = \{1, 2, ..., k\}$ where $k \to \infty$, then set $u = \frac{x}{k}$

-   obvious issue: $k \to \infty$? ($\Rightarrow$ choose $k$ based on desired level of accuracy)

#### Lecture 1 practical
```{r}
random_mcrng <- function(m, b, x0, n){
  x <- numeric(n+1)
  x[1] <- x0
  for (i in 2:n) {
    x[i]<- (b*x[i-1])%%m
  }
  u <- x/m
}

U<- random_mcrng(x0 = 207, m = 1307, b = 53, n=10000) 

hist(U)
```

The pseudo-randomly generated numbers u1, . . . , un are approximately evenly distributed over the interval [0, 1]. Also, the sample mean and sample variance of the generated numbers are very close to the theoretical
mean and variance of a U(0, 1)-distributed random variable, which are 1/2
and 1
12 = 0.083, respectively. This
suggests that the MCRNG algorithm that we implemented is doing a good job at generating U(0, 1)-distributed
random numbers.


## Simulation of other random variables

from book

### Bernoulli random variables

Write an R function which simulates the outcomes of a student guessing
at a True–False test consisting of n questions.
(a) Use the function to simulate one student guessing the answers to a
test with 10 questions; calculate the number of correct answers for
this student.
(b) Simulate the number of correct answers for a student who guesses
at a test with 1000 questions.

```{r}
set.seed(1234)
guessing <- function(success_probability = 0.5, n){
  guesses <- runif(n)
  correct_guess <- guesses<success_probability
  number_correct <- sum(correct_guess)
  
  return(number_correct)
  }

# test 
guessing(n=10)

#(b)
guessing(0.5, n=1000)
```

3 Write an R function which simulates 500 light bulbs, each of which has
probability 0.99 of working. Using simulation, estimate the expected
value and variance of the random variable X, which is 1 if the light bulb
works and 0 if the light bulb does not work. What are the theoretical
values?

```{r}
functional <- function(success_probability = 0.99, n){
  #x <- runif(n)
  #ifelse(x < success_probability, 1, 0)
  rbinom(n, size = 1, prob = success_probability)
}

X<- functional(n = 500)
sum(X)
mean(X)
var(X)

```

theoretical values: expectation per bulb = 0.99 and variance is = p*q (prob of success 
* prop of fail) = 0.99*0.01


### Binomial random variables

Compute the probability of getting four heads in six tosses of a fair coin
```{r}
dbinom(x = 4, size = 6, prob = 0.5)
```
1 Suppose the proportion defective is 0.15 for a manufacturing operation.
Simulate the number of defectives for each hour of a 24-hour period,
assuming 25 units are produced each hour. Check whether the number of
defectives ever exceeds 5. Repeat, assuming p = 0.2 and then p = 0.25

```{r}
no.defectives <- rbinom(n = 24, size= 25, prob = 0.15)
any(no.defectives>5)
no.defectives <- rbinom(24, 25, prob = 0.2)
any(no.defectives>5)
no.defectives <- rbinom(24, 25, prob = 0.25)
any(no.defectives>5)
```

Use simulation to estimate the mean and variance of a binomial random variable with n = 18 and p = 0.76. Compare with the theoretical
values.
```{r}
X <- rbinom(n = 1000, p = 0.76, size = 18)
mean(X); var(X)
```

theoretical
```{r}
#n*p
18*0.76
#npq
18*0.76*(1-0.76)

```

The following function simulates binomial pseudorandom numbers
by summing up the corresponding independent Bernoulli random
variables.
```{r}
#Generate binomial pseudorandom variables by summing Bernoulli
ranbin2 <- function(n, size, prob) {
  #'singlenumber' generates one binomial random variable.
  singlenumber <- function(size, prob) {
    x <- runif(size)
    N <- sum(x < prob)
    N
  }
  replicate(n, singlenumber(size, prob))
}

```

### Poisson random variables




