# Probability

@pishro-nik2014

## Random Variables

A random experiment is a procedure where uncertain events are observed. An **outcome** is the result of such an experiment, and the collection of all possible outcomes is called the **sample space** $S$. For example:

-   a coin toss: $S = \{heads,tails\}$

-   roll a 6 sided die (D6): $S = \{1, 2, 3, 4, 5, 6\}$

-   number of rainy days in June: $S = \{0, 1, ..., 30\}$

Repetitions of a random experiment is known as a **trial.** If a coin is tossed twice $S = \{(H, H), (T, H), (H, T), (T, T)\}$.

An **event** is a subset of the sample space.

The numerical results of an a random experiment are **random variables(RV)**. For the example of a coin tossed twice, numeral values can be assigned to each outcome: $(H, H) = 0, (T, H) = 1, ...$

The set of possible values for random variable $X$ is denoted by $Range(X)$, $R_X$, $\Omega$.

1.  I roll a D6 10 times. Let X be the number of 1's observed. Then $R_X = \{0, 1, ..., 10\}$

2.  How many attempts it takes for Manchester United (F.C.) defense to tighten up. Let Y represent the total number of games before they manage to keep a clean sheet. Then $R_Y = \{ 1,2, ...\} = \mathbb{N}$ i.e. Y can be any positive integer.

3.  The random variable T is the time (in hours) from now until the next earthquake occurs. This can be any non negative real number $R_T = [0, \infty)$

## Discrete random variables

A random variable X is called discrete if:

1.  $\Omega$ is a finite set. E.g. $\Omega_X = \{1, 2, 3, 4\}$.

2.  a countably infinite set. E.g.:

    -   Natural numbers $\mathbb{N} = \{1, 2, 3, ...\}$,

    -   Integers $\mathbb{Z} = \{..., -1, 0, 1,...\}$,

    -   **Rational number**, $\mathbb{Q}$ (number that can be expressed as a fraction), etc..

If $\Omega$ isn't a countable set, X is continuous.

### Definitions explained through an example

If a coin is tossed four times:

-   The sample space (S) ---the list of all ($2^4$) possible outcomes--- is $S = \{(H, H, H, H), (H, H, H, T), ....\}$.

-   We define the RV, $X$, as the number of heads that come up in the four coin tosses. Then the range of X is $\Omega_X = \{x_1 = 0, x_2 = 1,..., x_5= 4\}$.

-   An **event** defined as the set of outcomes ***s*** in the sample space ***S***. For example an event, **A**, could be "getting exactly 2 heads". Then **A** would include all outcomes from the sample space where there are 2 heads, e.g.: $(HHTT), (HTHT),$ etc. In general terms $A=\{s∈S|X(s)=x_k\}$

-   Probabilities of events $\{X=x_k\}$ are formally shown by the probability mass function (**pmf**) of $X$, i.e. showing the probability of each value of $\Omega_X$: $P(X=0), ..., P(X=4)$. Formally, the **pmf**: $f_X(x_k) = P(X=x_k)$.

```{r, echo=FALSE}

#| label: fig-coin_pmf1

#| fig-cap: "e.g. P(X = 1): The probability of getting 1 head (HTTT, THTT, TTHT, TTTH) = 4/(2^4) = 0.25." 

# number of trials (n) 
n <- 4
# probability of success (p) for a fair coin
p <- 0.5

# possible number of heads (x) and the corresponding probabilities (pmf)
coin_tosses <- data.frame(Number_of_Heads = 0:n,
                          Probability = dbinom(0:n, size = n, prob = p))

# PMF plot 

plot(coin_tosses$Number_of_Heads, coin_tosses$Probability, 
     type = "h", lwd = 3, 
     xlab = "x", ylab = "P(x)", 
     main = "PMF", ylim = c(0, 0.5))

```

-   To be a pmf, $f_X$ must have the following properties:

    -   $f_X \ge 0$ for $x \in \Omega$. i.e $P(X =x_k) < 0$

    -   $\Sigma_{x \in \Omega} f_X = 1$

    From our example, for in each case $f_X \ge 0$ and the sum is `r sum(coin_tosses$Probability)`

```{r, echo=FALSE}
kableExtra::kable(coin_tosses)
```

-   The **cumulative distribution function (cdf)** $F_X(x) = P(X ≤ x)$

```{r}
coin_tosses$cdf <- cumsum(coin_tosses$Probability)

plot(coin_tosses$Number_of_Heads, coin_tosses$cdf, 
     type = 's', bty = "l", 
     xlab = 'x', ylab = 'cumulative probability',
     main = 'Cumulative distribution function (CDF)')
abline(h = 1, lty = 2)
```

- The cdf FX is a non-decreasing function for which .... 


### Examples

#### Uniform Distribution

#### Bernoulli Distribution

#### Binomial distribution

``` {webr-r}
n = 5; p = 0.5
omega = 0:n
pmf = dbinom(x = omega, size = n, prob = p)
bin.distr = data.frame(x = omega, pmf)
bin.distr
```

#### **Poisson Distribution**

#### Geometric Distribution

#### Negative Binomial (Pascal) Distribution

#### Hypergeometric Distribution
